The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT

### 2012-09-05 talk/vmil2012/paper.tex
Add a paragraph about the violation of immutability.
reviewers' comments

@@

@@

Our aim is to help understand the constraints when implementing guards
and to describe the concrete techniques used in the various layers of RPython's
tracing JIT. All design decisions are be motivated by an empirical analysis of the
frequency and the overhead related to guards.

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations about guards and
describe based on them the reasoning behind their implementation in
RPython's tracing just-in-time compiler. The contributions of this paper are:
\begin{itemize}
  \item An analysis guards in the context of RPython's tracing JIT,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of RPython's JIT and a description of the rationale behind the design.
\end{itemize}

@@

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language is a
statically typed object-oriented high-level subset of Python. The subset is chosen in such a way to make type inference possible\cite{ancona_rpython:_2007}.
The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing is a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs are typically
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

@@

@@

@@

@@

@@

@@

@@

%\section*{Appendix}
%\todo{remove this section and the figures}
%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
%\begin{figure*}
%\centering
%\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
%\caption{Relative frequency of operations before and after optimization}
%\label{fig:ops_pie}
%\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
%\listoftodos
\end{document}

>>>>

\newcommand{\bigtodo}[1]{\todo[inline,color=red!30]{#1}}

@@

\bigtodo{reviewer c: The use of an interpreter is a typical choice, but it is
not the only choice (e.g., Jikes RVM is compile only) so it might make
sense to introduce this idea earlier.}

@@

Our aim is to help understand the constraints when implementing guards
and to describe the concrete techniques used in the various layers of RPython's
tracing JIT. All design decisions are motivated by an empirical analysis of the
frequency and the overhead related to guards.

@@

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations about guards and
describe based on them the reasoning behind their implementation in
RPython's tracing just-in-time compiler. The contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of RPython's tracing JIT,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of RPython's JIT and a description of the rationale behind the design.
\end{itemize}

@@

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language is a
statically typed object-oriented high-level subset of Python. The subset is chosen in such a way to make type inference possible\cite{ancona_rpython:_2007}.
The language provides
several features such as automatic memory management
and just-in-time compilation.\todo{reviewer d: I argue that jit compilation is
not a feature of a language, but a feature of a language
\emph{implementation}.}
When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

\subsection{RPython's Tracing JIT Compiler}
\label{sub:tracing}

\bigtodo{reviewer c: In 2.2 there was a comment that guards can be used to
specialize based on type in dynamic languages. Is this only applicable to
dynamic languages, or could the same thing occur due to simple inheritance?}

Tracing is a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs are typically
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

@@

\bigtodo{reviewer b: sadly no graphs}

\bigtodo{reviewer b: One missing data set in the empirical study
is excution time performance info for the 
various guard implementation choices.
The empirical data focuses on space not
time.}

\bigtodo{reviewer c: I found the evaluation less detailed than I was hoping
for, which was a shame as I was very interested in learning how well everything
performed after reading the earlier sections!}

\bigtodo{reviewer c:
I would have liked to see an evaluation based on execution time rather than
operations. I do not have a good sense of the relative cost of different
operations, and from my experience know that it can be misleading to use such a
measure.}

\bigtodo{reviewer c:
For all of the results some aggregate values would be useful (max/min, etc.)
}

@@

\bigtodo{reviewer c: Figure 7 was printed out of order. It was not clear if
    Figure 7 was talking about static measures (count once per compilation),
    vs. dynamic measures (count once per actual execution). When I have done
    studies like this in the past the difference between these two measures can
    be very interesting. One has more of an impact on memory and compiler
    issues, while the other will more significantly affect execution
    performance.
}

@@

\bigtodo{reviewer c: It might be possible to use a chart for
Figure~\ref{fig:failing_guards} to give more information?}

\bigtodo{reviewer d:
I would also like to see the number of guard that account for 99\% (or even
99.9\%) of all guard failures, in addition to the 50\% number.}

@@

\bigtodo{ reviewer c: I found it difficult to understand how the size of the
resume structures are approximated, or what you mean in the footnote where you
say it is reconstructed from log files. }

@@

\bigtodo{reviewer c: In 6.1 you claim that the elimination of snapshots is
orthogonal to resume data compression. While they are techniques that can
be used individually or together, I think there may be non-trivial
interactions, in particular eliminating snapshots may alter opportunities
for compression.}

@@

\bigtodo{reviewer d: You state that Kotzmann et al. do not explain how the
debug information is store in a compact way. They have a second paper that
describes exactly that: Thomas Kotzmann, Hanspeter Mössenböck: Run-Time
Support for Optimizations Based on Escape Analysis Proceedings of the
International Symposium on Code Generation and Optimization, San Jose,
March 2007.}

@@

%\section*{Appendix}
%\todo{remove this section and the figures}
%\begin{figure*}
%    \include{figures/ops_count_table}
%    \caption{Relative numbers of operations in the traces generated for
%    different benchmarks}
%    \label{fig:ops_count}
%\end{figure*}
%\begin{figure*}
%\centering
%\includegraphics[width=\textwidth]{figures/ops_pie.pdf}
%\caption{Relative frequency of operations before and after optimization}
%\label{fig:ops_pie}
%\end{figure*}
\bibliographystyle{abbrv}
\bibliography{zotero,paper}
\listoftodos
\end{document}
